### 2011-11-17 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

Specifically, it is desired to determine the XML Schema built-in types \todo{link do appendixu} of expressions, if it is possible. Also, for each path expression we want to determine if it selects elements or attributes and we want to store the entire path in the type information. This information can be used for example in a $FLWOR$ expression \todo{priklad nejakeho FLWOR a ukazat ako vieme zistit, ktore elementy su prechadzane}. \todo[inline]{Uistit sa, ze takto nam to staci.}

To capture sequences, we assign each type with its cardinality as proposed in \todo{ref na Schajbalovu diplomku}. The cardinality expresses one of the following five sequence types.

\begin{itemize}
\item An empty sequence.
\item A sequence of exactly one item.
\item A sequence containing zero or one item.
\item A sequence containing zero or more items.
\item A sequence containing one or more items.
\end{itemize}

@@

@@

@@

\STATE $type = null$
\IF{$exprNode$ represents a literal value}
     \STATE $type =$ particular literal value type
\ELSIF{$exprNode$ represents a call of function $f$}
     \STATE $type = getFunctionReturnType(f)$
\ELSIF{$exprNode$ represents variable reference $var$}
     \STATE $type = getVariableType(contextVarTypes, var)$
\ELSIF{$exprNode$ represents path expression $path$}
     \STATE $type = TODO$
\ELSIF{$exprNode$ represents operator $op$}
     \STATE TODO
\ELSIF{$exprNode$ represents $FLWOR$ construct $flwor$}
     \STATE  $type =$ type of an expression in the return clause of $flwor$
\ENDIF

@@

The handling of new variable definitions relies on the fact that the definitions make the new variables valid only in nodes with a higher sequence number when numbered in the post-order \todo{overit}. Therefore, the left-most subnode of a certain node can be processed without an extension of the variable context. And, if the left-most subnode extends the variable context for the following nodes in the post-order numbering, it can be easily handled because the types of the new variables can be directly determined since the binding expressions are already processed (thus, their types are known). Every node "knows" wheather or not it may define new variables for nodes with higher sequence numbers. This is expressed in the algorithm by the condition at line 10. This condition is double-checked at line 3 where it meaning is that every node even "knows" if its particular subnode may define new variables. An example is $FLWOR$ construct. The processing of $FLWOR$ knows that the first part (the first subnode) of this node may bind new variables using $for$ and $let$ constructs. After the subnode is processed, the processing of the $FLWOR$ node looks if any new variables were bound and if so the context of local variables is extended. This is done at line 4. Function $mergeContextVarTypes$ presented in \ref{ALG_function_mergeContextVarTypes} writes every record from its second argument (the new variables) into ist first argument (the variable context). If there are records for variables with the same names in the context, they are overwritten to correspond to the variable overlaping. It is important to note that we assume the variable context is not passed by refence but by value, and hence, the variable context of a certain node is not affected by the recursive processing of its subnodes.

@@

\todo[inline]{Popisat odvodzovanie z volania funkcie}
\subsection{Function Calls}
This case is quite simple. The algorithm encounters a function call and one of the arguments is a known set of elements of attributes (subexpression $E$ representing $S$). To determine the type of $S$, it is only needed to determine the type of the corresponding formal argument from the definition of the function. The function is either a built-in one so its definition is known or it is defined in the prolog section. External functions are not processed as was mentioned already.

>>>>

We want to be able to capture the following types.

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Type $xs:untyped$.
\item Type representing any item. \todo{potrebujeme?}
\item Type representing an element, attribute, item, node, text node, comment, processing instruction, document node.
\item Type representing a set of elements or attributes selected by a certain path expression. The path expression is included in this type.
\end{itemize}

\todo[inline]{Akeho typu je napr //(elementA * 2)?}

To capture sequences, we assign each type with its cardinality as proposed in \todo{ref na Schajbalovu diplomku}. Every type can be perceived as a sequence. A type representing one value or one item can be perceived as a sequence of exactly one item. The cardinality expresses one of the following five sequence types.

\begin{itemize}
\item An empty sequence.
\item A sequence of exactly one item.
\item A sequence containing zero or one item (modifier $?$).
\item A sequence containing zero or more items (modifier $*$).
\item A sequence containing one or more items (modifier $+$).
\end{itemize}

@@

\todo[inline]{Este by sa dala urobit analyza volania funkcii, do ktorych su predavane nejake elementy/atributy o ktorych vieme ich path expression. Vysledkom tejto funkcie by totiz zas mohli byt elementy/atributy o ktorych vieme (pozmenenu) path expression a nie len nezname elementy/atributy.}

@@

\todo[inline]{Udrzat si referencie na definicie funkcii pre dalsie pouzitie.}

@@

\STATE $type = null$
\IF{$exprNode$ represents a literal value}
     \STATE $type =$ particular literal value type, exactly-once cardinality
\ELSIF{$exprNode$ represents a call of function $f$}
     \STATE $type = getFunctionReturnType(f)$
\ELSIF{$exprNode$ represents variable reference $var$}
     \STATE $type = getVariableType(contextVarTypes, var)$
\ELSIF{$exprNode$ represents path expression $path$}
     \STATE $type = TODO$
\ELSIF{$exprNode$ represents operator $op$}
     \STATE TODO
\ELSIF{$exprNode$ represents $FLWOR$ construct $flwor$}
     \STATE  $type =$ type of an expression in the return clause of $flwor$
\ENDIF

@@

The handling of new variable definitions relies on the fact that the definitions make the new variables valid only in nodes with a higher sequence number when numbered in the post-order \todo{overit}. Therefore, the left-most subnode of a certain node can be processed without an extension of the variable context. And, if the left-most subnode extends the variable context for the following nodes in the post-order numbering, it can be easily handled because the types of the new variables can be directly determined since the binding expressions are already processed (thus, their types are known). Every node ''knows'' wheather or not it may define new variables for nodes with higher sequence numbers. This is expressed in the algorithm by the condition at line 10. This condition is double-checked at line 3 where it meaning is that every node even ''knows'' if its particular subnode may define new variables. An example is $FLWOR$ construct. The processing of $FLWOR$ knows that the first part (the first subnode) of this node may bind new variables using $for$ and $let$ constructs. After the subnode is processed, the processing of the $FLWOR$ node looks if any new variables were bound and if so the context of local variables is extended. This is done at line 4. Function $mergeContextVarTypes$ presented in \ref{ALG_function_mergeContextVarTypes} writes every record from its second argument (the new variables) into ist first argument (the variable context). If there are records for variables with the same names in the context, they are overwritten to correspond to the variable overlaping. It is important to note that we assume the variable context is not passed by refence but by value, and hence, the variable context of a certain node is not affected by the recursive processing of its subnodes.

@@

\subsection{Function Calls}
This case is quite simple. The algorithm encounters a function call and one of the arguments is a known set of elements of attributes (subexpression $E$ representing $S$). To determine the type of $S$, it is only needed to determine the type of the corresponding formal argument from the definition of the function. The function is either a built-in one so its definition is known or it is defined in the prolog section. External functions are not processed as was mentioned already.
